Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking

### 2023-08-02 manuscript/manuscript.qmd
update README.md

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "Studying sleep is vital in health research, but the gold standard, polysomnography, is costly and impractical for large-scale studies. An affordable alternative is using wearable accelerometers. While wrist and hip-worn devices are commonly used in sleep research, thigh-worn accelerometers have been relatively unexplored. Our study evaluated machine learning and deep learning models utilizing data from thigh-worn accelerometers to estimate sleep and sleep quality metrics, comparing them with an EEG-based sleep monitor. The dataset consisted of data from 585 days and nights, comprising accelerometry and EEG-based sleep estimates from children aged 6-10. We employed both sequential and multiclass model strategies on both raw and filtered data. The most effective model was XGBoost, which performed well when applied to 5-minute median filtered data, exhibiting small biases in sleep period time (0.2 minutes), total sleep time (-7 minutes), sleep efficiency (-1.1%), and wake after sleep onset (-0.9 minutes). Furthermore, the XGBoost model showed a robust correlation (0.66, 95% CI: 0.61 - 0.7) with total sleep time, indicating its potential. However, despite these favorable results in bias, our study revealed large limits of agreements in accordance with previous research on hip- and wrist-worn devices. In conclusion, we present promising results in using machine learning technieques to estimate sleep quality metrics, however, accurately classifying awake periods during in-bed time remained challenging. Moreover, additional improvements are necessary to precisely assess individual sleep quality metrics due to the notable limits of agreement."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
on 5-minute median filtered ZM predictions, and the 5-minute
median-smoothed ZM-derived sleep quality metrics. The Bland-Altman plot
for the SPT and TST reveals a good level of agreement with the ZM, as
evidenced by a bias close to zero. Interestingly, a portion of the data
points are located near the zero line indicating perfect agreement. The
scatterplot for SPT also demonstrates a positive trend, indicating a
moderate linear correlation between the XGBoost model and the ZM-derived
sleep quality metrics. The bias and LOA for TST are comparable to those
observed for SPT, indicating a consistent level of agreement between the
two methods. The scatterplot for TST also shows a slightly higher
correlation, primarily driven by the absence of extreme
outliers.Furthermore, the remaining three sleep quality metrics, SE,
LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST.
~~This outcome is expected as achieving 100% sleep efficiency is
relatively rare, resulting in less disagreement between the methods as
values approach the upper limit. Conversely, as sleep efficiency
decreases, the potential for discrepancies and differing interpretations
between the methods increases, leading to greater heteroscedasticity.~~
A moderate positive linear correlation is observed between the XGBoost
model and ZM-derived sleep quality metrics for SE, however, a poor
correlation is observed for LPS and WASO.

>>>>

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
shorttitle: "Improving Sleep Quality Estimation"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "Studying sleep is vital in health research, but the gold standard, polysomnography, is costly and impractical for large-scale studies. An affordable alternative is using wearable accelerometers. While wrist and hip-worn devices are commonly used in sleep research, thigh-worn accelerometers have been relatively unexplored. Our study evaluated machine learning and deep learning models utilizing data from thigh-worn accelerometers to estimate sleep and sleep quality metrics, comparing them with an EEG-based sleep monitor. The dataset consisted of data from 585 days and nights, comprising accelerometry and EEG-based sleep estimates from children aged 6-10. We employed both sequential and multiclass model strategies on both raw and filtered data. The most effective model was XGBoost, which performed well when applied to 5-minute median filtered data, exhibiting small biases in sleep period time (0.2 minutes), total sleep time (-7 minutes), sleep efficiency (-1.1%), and wake after sleep onset (-0.9 minutes). Furthermore, the XGBoost model showed a robust correlation (0.66, 95% CI: 0.61 - 0.7) with total sleep time, indicating its potential. However, despite these favorable results in bias, our study revealed large limits of agreements in accordance with previous research on hip- and wrist-worn devices. In conclusion, we present promising results in using machine learning technieques to estimate sleep quality metrics, however, accurately classifying awake periods during in-bed time remained challenging. Moreover, additional improvements are necessary to precisely assess individual sleep quality metrics due to the notable limits of agreement."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
format:
  elsevier-pdf:
    keep-tex: false
    # fig-pos: 'b'
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
  simple-article-pdf:
    indent: false
    classoption: 
     - 9pt
     - twoside
  docx:
    number-sections: true
  html: 
    csl: https://www.zotero.org/styles/nature
pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
on 5-minute median filtered ZM predictions, and the 5-minute
median-smoothed ZM-derived sleep quality metrics. The Bland-Altman plot
for the SPT and TST reveals a good level of agreement with the ZM, as
evidenced by a bias close to zero. Interestingly, a portion of the data
points are located near the zero line indicating perfect agreement. The
scatterplot for SPT also demonstrates a positive trend, indicating a
moderate linear correlation between the XGBoost model and the ZM-derived
sleep quality metrics. The bias and LOA for TST are comparable to those
observed for SPT, indicating a consistent level of agreement between the
two methods. The scatterplot for TST also shows a slightly higher
correlation, primarily driven by the absence of extreme
outliers.Furthermore, the remaining three sleep quality metrics, SE,
LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST. A
moderate positive linear correlation is observed between the XGBoost
model and ZM-derived sleep quality metrics for SE, however, a poor
correlation is observed for LPS and WASO.
